---
title: "Quality of Life Dataset Analysis"
format: 
  revealjs:
    theme: blood
    fontsize: 17pt
editor: visual
---

```{r}
#| include: false
library(ggplot2)
library(tidyverse)
library(maps)
library(mapproj)
library(RColorBrewer)
library(ggridges)
library(countrycode)
library(gridExtra)
load("C:/Users/msidd/Downloads/prjdata.RData")
```

## Introduction

-   **Key Idea:** Quality of life is shaped by trade-offs between safety, affordability, environmental conditions, and access to services.

<!-- -->

-   **Objective:** Investigate how economic prosperity influences well-being through safety, cost of living, pollution, commute times, and healthcare.

<!-- -->

-   **Data Source:** Numbeo – a global database of living conditions.

<!-- -->

-   **Main Question:** Does more money mean a better life?

## Research Questions

-   **Safety & Cost of Living:** Do safer countries have a higher cost of living?

-   **Pollution & Commute Times:** How does economic growth impact environmental quality and transportation?

-   **Healthcare & Purchasing Power:** Does greater wealth lead to better healthcare access?

## **Correlation between Safety Value and Cost of Living?**

-   **Cost of Living vs Safety Value**
    -   A negative correlation—safer countries tend to have a higher cost of living.
-   Wealthier nations invest more in security, infrastructure, and services.
-   Economic growth improves safety but raises living costs.

```{r}
#| warning: false
qol$Cost.of.Living.Category = factor(qol$Cost.of.Living.Category, 
                                    levels = c("Very Low", "Low", "Moderate", 
                                                 "High", "Very High"))
qol_clean = qol |> 
  filter(!is.na(Safety.Value) & !is.na(Cost.of.Living.Category))

# Graph 3: Cost of Living vs Safety Value
ggplot(qol_clean, aes(x = Cost.of.Living.Category, y = Safety.Value,
                     fill = Cost.of.Living.Category)) +
  # 1. Violin Plot (with modifications)
  geom_violin(alpha = 0.6, draw_quantiles = c(0.5)) +  # Violin plot, transparent, median line
  # 3. Aesthetics
  scale_fill_brewer(palette = 'RdYlBu') +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title.x = element_text(size = 12, face = "bold"),
        axis.title.y = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10)) +
  labs(y = "Safety Value", x = "Cost of Living Category",
       title = "Distribution of Safety Value by Cost of Living Category")
```

## Do pollution levels and commute times relate?

-   **Pollution Value by Traffic Commute Time Category**
    -   Longer commute times are associated with higher pollution levels.
-   Implication: Traffic congestion is a major contributor to environmental decline.

<!-- -->

-   Solution: Investment in sustainable transport is essential.

```{r}
qol$Traffic.Commute.Time.Category = factor(x = qol$Traffic.Commute.Time.Category, levels = c("Very Low", "Low", "Moderate", "High", "Very High"))
ggplot(data = qol |> filter(!is.na(Traffic.Commute.Time.Category), Pollution.Value > 0), aes(x = Traffic.Commute.Time.Category, y = Pollution.Value)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", outlier.shape = 8, outlier.colour = "red", outlier.size = 2) +
  labs(
    title = "Pollution Value by Traffic Commute Time Category",
    x = "Traffic Commute Time Category",
    y = "Pollution Value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold")
  )
```

## Does higher purchasing power lead to better healthcare access?

-   **Purchasing Power vs Health Care Quality**
    -   Strong positive correlation: wealthier nations tend to have better healthcare
-   Outliers - Angola, Timor-Leste, and Kiribati:
    -   Due to their unusually high purchasing power - suggesting that factors other then wealth, such as geographical isolation could be limiting the impact of economic prosperity in healthcare quality

```{r}
#| message: false
#| warning: false
top_3_outliers = qol |> 
  filter(Health.Care.Value > 0 & Purchasing.Power.Value > 0) |> 
  arrange(desc(Purchasing.Power.Value)) |> 
  head(3)

# Scatter plot with top 3 outliers labeled on the y-axis
ggplot(qol |> filter(Health.Care.Value > 0 & Purchasing.Power.Value > 0), aes(x = Health.Care.Value, y = Purchasing.Power.Value)) +
  geom_point(color = "blue", alpha = 0.6, size = 3) +  
  geom_smooth(method = "lm", color = "red", size = 3, se = FALSE) + 
  geom_point(data = top_3_outliers, aes(x = Health.Care.Value, y = Purchasing.Power.Value), color = "red", size = 4) +  # Top 3 outliers in red
  geom_text(data = top_3_outliers, aes(label = country), vjust = -1, color = "black", size = 3) +  # Label top 3 outliers
  labs(
    title = "Purchasing Power vs Health Care Quality",
    x = "Health Care Quality Value",
    y = "Purchasing Power Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 12, face = "bold")
  )
```

## Conclusion: The Trade-Offs of Modern Living

-   Economic strength enhances safety, healthcare, and infrastructure—but at a cost.

<!-- -->

-   Pollution and long commutes are byproducts of economic progress.

<!-- -->

-   **There is no perfect place to live—countries and individuals must navigate trade-offs.**

<!-- -->

-   **Policy Recommendations:**

    -   Balance safety investments with affordability.

    -   Prioritize sustainable urban planning to reduce pollution.

    -   Strengthen healthcare accessibility beyond just economic growth.
